---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
========================================================

##Loading and preprocessing the data
```{r LoadPreprocess, echo = TRUE}
library(knitr);
rawdata <- read.csv("activity.csv")
adata <- na.omit(rawdata)
adata$date <- as.factor(as.character(adata$date)) # Removes date factors corresponding to NA values. 
```
##What is mean total number of steps taken per day?
```{r libdplyr, echo = TRUE,warning=FALSE}
library(dplyr)
```
```{r calcPlotSteps, echo = TRUE}
dq1a <- adata %.%
        group_by(date) %.%
        summarize(totalsteps = sum(steps, na.rm = TRUE))

# Alternative Code for dq1a using TApply
# dqa <- tapply(adata$steps, adata$date, sum, na.rm = TRUE);
# dq1a <- cbind.data.frame(date = names(dqa),totalsteps = unname(dqa))

hist(dq1a[,2],
        xlab = "Total Steps Per Day",
	ylab = "Number of Days",
	main = "Frequency of Total Steps in a day",
        col = "red")

meansteps <- mean(dq1a$totalsteps, na.rm = TRUE)
mediansteps <- median(dq1a$totalsteps, na.rm = TRUE)

print(sprintf("Mean total steps taken per day: %f ", meansteps))
print(sprintf("Median total steps taken per day: %f ", mediansteps))
```

##What is the average daily activity pattern ?
```{r dailypattern, fig.width = 7, fig.height=6, echo=TRUE}
dq2a <- adata %.%
        group_by(interval) %.%
	summarize(avgsteps = mean(steps))

# Alternative method to calculate dq2a
# dqa <- tapply(adata$steps, adata$interval, mean,na.rm = TRUE) 
# dq2a <- cbind.data.frame(interval = as.integer(names(dqa)),avgsteps = unname(dqa))

plot(dq2a$interval, 
	dq2a$avgsteps, 
	type = "l",
	xlab = "5-min Interval",
	ylab = "Average Num of Steps",
	main = "Average Daily Activity Pattern")

maxavginterval = dq2a$interval[dq2a$avgsteps == max(dq2a$avgsteps)]

print(sprintf("5-min time Interval with maximum average steps taken per day: %i ", maxavginterval))
```

##Imputing Missing Values
```{r imputemissingvals, echo=TRUE }
nacount <- nrow(rawdata[!complete.cases(rawdata),])
print(sprintf("Row count with NA data: %i ", nacount))
```
###     Strategy:
###Replace NA step values for corresponding intervals with mean interval values calculated above
```{r newdataset, echo=TRUE}
newdata <- rawdata
sapply(unique(rawdata$interval),
		function(x)
		newdata[!complete.cases(newdata) & (newdata$interval == x),1] <<- dq2a$avgsteps[dq2a$interval == x])
```
### Plot Histogram of total steps with imputed data
```{r newdatahist, echo=TRUE}
dqc <- tapply(newdata$steps, newdata$date, sum, na.rm = TRUE);
dq1c <- cbind.data.frame(date = names(dqc),totalsteps = unname(dqc))

hist(dq1c[,2],
	xlab = "Total Steps Per Day",
	ylab = "Number of Days",
	main = "Frequency of Total Steps in a day",
        col = "blue")

meansteps <- mean(dq1c$totalsteps)
mediansteps <- median(dq1c$totalsteps)

print(sprintf("Mean total steps taken per day: %f ", meansteps))
print(sprintf("Median total steps taken per day: %f ", mediansteps))
```
##Are there differences in activity patterns between weekdays and weekends?
```{r compareweekends, echo=TRUE}
newdata$date <- as.Date(newdata$date, "%Y-%m-%d")
newdata$weekend <- "weekday"
newdata$weekend[weekdays(newdata$date) %in% c("Saturday","Sunday")] <- "weekend"
newdata$weekend <- as.factor(newdata$weekend)

dq2a <- newdata %.%
		group_by(interval, weekend) %.%
		summarize(avgsteps = mean(steps))

library(ggplot2)
qplot(x = interval,
	y = avgsteps,
	data = dq2a,
	geom = c("line"),
	facets = weekend~.,
	ylab = "Average number of Steps")
```